Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2016-01-12 tmm.tex
Add more discussion on the robustness of the controller
Change the first half part according to reply of editing service

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenge brought by bandwidth fluctuation and improve the experience of watching online videos, an adaptive video streaming system which can adjust video quality according to actual network condition is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and verified for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. On that basis, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enables R-D optimized bitstream extraction under certain bitrate constraint. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine the suitable bitrate for transmission. By monitoring and predicting bandwidth information of the past, the current and the future, the PID-based quality control algorithm is able to reduce quality fluctuation while still keeping a high quality level. Experimental results show that, comparing with the baseline software, the proposed system integrating the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 1.24 to 0.69, and at the same time deliver higher video quality, with the PSNR average increased by 0.83 dB.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
\IEEEPARstart{V}{ideo} streaming applications are becoming more and more popular on the Internet. Examples of the most common usage scenarios include Video-on-Demand (VoD), live broadcasting, video-based online education, etc. In these applications, video data is constantly delivered to different clients over various transmission channels, and the client media player can begin to play the video before the entire file has been transmitted. Among other features, video streaming makes it easy to watch video anywhere and anytime, which represents the concept of Ubiquitous Multimedia Access (UMA), a very common topic within the multimedia community during the past few years \cite{Gualdi08}. Due to its convenience, video streaming also attracts a large number of users and becomes an increasingly important way for people to consume video content \cite{Chen13}.

In video streaming, the network condition significantly influences the user's watching experience. To provide better network condition, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm named OpenFlow, noticeable results are achieved. For instance, a solution for improving network throughput in hybrid network is proposed in \cite{Li14} and a demonstration of OpenFlow-controlled network orchestration for video manycast is present in \cite{Xue15}. Although these studies have offered a promising way to optimize network condition, currently traditional best-effort networks still dominate, and the variety of users' network environment remains to be one of the biggest challenges for video streaming. A user might use different types of network (Ethernet, Wi-Fi, Cellular, etc.) and might also switch from one type to another on going (e.g., from Wi-Fi to Cellular). Even in the same network, the end-to-end bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable, and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a fixed bitrate. To cope with this challenge, the video streaming system should be able to adjust the video's bitrate or quality according to the network condition. In other words, the video streaming system needs to be adaptive.

There are generally two options to enable adaptive video streaming. The first option is called simulcast, where several independent bitstreams, each with different video quality, are stored at the server, and the client selects one of them according to currently available bandwidth. Dynamic Adaptive Streaming over HTTP (DASH) \cite{DASH}, which is standardized by the Moving Picture Expert Group (MPEG), belongs to this option. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, an extension of the H.264/AVC video coding standard \cite{SVCOverview}, which allows scalability of video quality and bitrate with only one single bitstream. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate video quality and bitrate as required.

Due to the easy deployment of simulcast systems \cite{Bouten14}, they are already used by many commercial video websites (e.g., YouTube \cite{YouTube}) to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams require a lot of computation time and disk space. In order to provide a wide range of seamless quality adaptation at reasonable time and space cost, the adoption of SVC is necessary. An SVC video bitstream can be extracted at the fine granularity of one NALU (a very small partition of data), and performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast where the same content is encoded multiple times. Therefore, to adopt SVC in the video streaming framework is technically a better option and has attracted much attention in the academia \cite{Chuah12}-\cite{Cicalo14}.

While these studies mostly focus on Quality of Service (QoS) optimization and resource allocation for specific networks, there are two key problems that need to be primarily addressed in the SVC adaptive video streaming system. First, as the prerequisite of bitrate adaption, the whole bitstream needs to be extracted so that sub-streams with various bitrates can be obtained. This is called the bitstream extraction problem. Optimal bitstream extraction algorithm aims to extract a sub-stream with best video quality, or minimal distortion, under the constraint of a given bitrate. Second, when the bandwidth changes with time, it is necessary to make decisions about whether and how to adjust the video quality or bitrate. This problem is called quality control, which means controlling the video quality (or the bitrate, from the transmission point of view) to follow the change of the bandwidth. A good quality control algorithm should determine a suitable bitrate according to current network condition, and keep the video quality both smooth and at a high level.

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia contents in practice, is chosen as the basis of the proposed system. The original DSS project is modified and SVC support is added to it, for the purpose of both research and application. Then novel algorithms for bitstream extraction and quality control are designed and implemented to enable better quality of service for the user. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective Linear Error Model (LEM) is proposed and verified, which can be used to accurately estimate the distortion caused by discarding any combination of data packets from an SVC bitstream.
\item Second, utilizing distortion estimation with LEM, a greedy-like algorithm is designed to assign priority for each data packet according to its Rate-Distortion (R-D) impact, thus enables optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) controller is proposed to replace the native one in DSS, and much lower quality fluctuation is achieved.
\end{itemize}

@@

The rest of this paper is organized as follows. In section \ref{sec:analysis}, we analyze the problems in detail and introduce some related work on them. Then in section \ref{sec:bitstream-extraction}, the linear error model and the solution for the bitstream extraction problem are present. After that, the PID-based quality control algorithm is proposed in section \ref{sec:quality-control}. Section \ref{sec:experiment} lists the experimental results and compares our proposed solution with others. Finally, in section \ref{sec:conclusion}, we conclude this paper's work and point out its promising applications and extensions in the future.

@@

In this section, we analyze the two most concerned problems in SVC adaptive video streaming, i.e., bitstream extraction and quality control, and introduce some related work on them.

@@

Bitstream extraction is the operation of choosing a subset of data packets from a whole SVC bitstream to form a sub-stream. Fig. \ref{fig:Bitstream-Extraction} illustrates the structure of an SVC bitstream and a possible way to extract a sub-stream from it. The data packets in the bitstream are divided into different layers. The temporal layer (T0$\sim$T3) reflects the reference or dependency relationship between the frames. The arrows on the top point from the referenced frame to the frame which is dependent on it. The quality layer (Q0$\sim$Q2) indicates scalability of video quality within each frame. Data packets in the higher quality layers (i.e., Q1 and Q2) are partially discarded to implement the extraction. For example, the dash line demonstrates the extraction which will keep the data packets below the line and discard all the packets above it.

@@

As we can see, in its nature SVC bitstream extraction is a combinatorial optimization problem. There are many combinations of packets to be discarded, and we want to choose a combination that maximizes the video quality of the extracted sub-stream, under the constraint of a given bitrate. More specifically, this problem is in the form of ``0-1 Knapsack Problem" \cite{Knapsack}. Each data packet can be treated as an object with certain weight (the packet's size) and value (its contribution to the video quality). The solving process is mainly to decide whether or not a packet should be included in the final extracted sub-stream, or in the term of ``0-1 Knapsack Problem", to choose which object to pack up.

For a typical ``0-1 Knapsack Problem", the optimal solution can generally be found using the dynamic programming algorithm. However, this is not operational for the bitstream extraction problem, mainly due to the dependence between the data packets. First, the subset of packets to be extracted cannot be chosen arbitrarily, since if a packet is included, all the packets it depends on must also be included. For example, in Fig. \ref{fig:Bitstream-Extraction}, we cannot include the Q2 packet of a frame without including the Q1 packet of the same frame. Second, the value of a packet is not obvious and certain, since a packet's contribution to video quality depends on other packets and may vary in different extractions. There is no way to exactly tell the video quality of an extracted sub-stream, unless we perform the extraction and calculate the distortion for real. In fact, to the best of our knowledge, no method presented in the literature claims to find the optimal solution of the SVC bitstream extraction problem. All the effort is made to obtain a solution that is as good as possible.

In \cite{Amonou07}, Amonou et al. proposed a bitstream extraction framework based on the concept of \textit{Quality Layers}. This framework has been adopted by the SVC reference software, i.e., JSVM \cite{JSVM}, and achieves better R-D performance compared with the basic extractor of JSVM. In Amonou's framework, a Quality Layer value, which reflects a data packet's R-D impact, is calculated and used for R-D optimized bitstream extraction. The process of obtaining distortion impact of every packet is extremely time consuming, and the distortion estimation accuracy can also be further improved. So other distortion/error models were proposed to either reduce the computational complexity or improve the estimation accuracy. Sun et al. \cite{Sun09} and Maani et al. \cite{Maani09} constructed models to calculate distortion based on the drift propagation between frames. Sun et al.'s method is almost as good as JSVM in regard of performance, but has much reduced complexity. Maani et al.'s model can achieve better estimation accuracy than JSVM; however, since it is a training-based model, more computation is needed to get robust model parameters. Ramanathan et al. \cite{Ramanathan12} proposed their concept and calculation method of the ``Quality Metadata'' to signal and use in bitstream extraction. Yang et al. \cite{Yang13} proposed a method based on simulated annealing to solve the bitstream extraction problem. In both methods, the accuracy of distortion estimation affects the performance a lot. With a packet's distortion impact accurately estimated, its prioritization order during extraction can be determined. Then a priority value can be assigned to it and stored in the bitstream for optimized extraction in the future. In \cite{Lim06}, Lim et al. treated SVC bitstream as a tree structure and assigned extraction priority accordingly. In \cite{Zhu11}, Zhu et al. chose to assign priority values derived using the Langrange multiplier method.

In our previous work \cite{Zhang12}, we proposed a linear error model for distortion estimation and priority assignment. Good bitstream extraction performance is achieved based on that. However, this work assumes that the original video sequence is available for reference, which is not always true. How to handle the special characteristic of the no-reference case is lacking in the previous work and included in this paper.

@@

Quality control is the second problem we will address in this paper. The result of a quality control algorithm gives the suitable bitrate for transmission, which is directly related to the change of the video quality delivered to the user. In an SVC streaming system, the bitrate determined by the quality control algorithm is used as the constraint under which bitstream extraction is performed. Even in non-SVC streaming systems (DASH, for example), quality control is an important part that requires careful design.

For a quality control algorithm, there are three targets to be achieved. The first is to ensure the continuity of video playing. In real-time streaming system, each data packet has its own playtime, which is also regarded as the deadline of the transmission. If the packet does not arrive at client side before the deadline, the media player will be paused due to lack of data, and that should be avoided. The second is to keep a smooth video quality. Although network conditions may change rapidly, it is necessary for the streamed video to maintain a relatively stable quality level, since frequent change of video quality will lead to poor user experience. The third target is to make the video quality as good as possible. In other words, the available bandwidth should be fully utilized. Suppose we keep sending video data in a low bitrate (with low video quality as well), neither the playback will pause nor any quality fluctuation will happen. However, a low video quality is certainly not what the user want.

The state-of-the-art solutions to avoid playback pause in client media player mainly focus on how to select and schedule the video packets. In \cite{Gao06}, K. Gao et al. proposed a priority based scheduling strategy which guarantees packets of higher importance will be transmitted earlier. Among the packets with the same priority, those having the earliest deadline will be transmitted first. A similar solution is proposed in \cite{Schierl10}, where video data is pre-buffered in several buffers of different priorities. Whenever re-buffering occurs, quality will be degraded to a lower level to meet the time and rate constraint. Although the priority based scheduling strategies perform well in maintaining a non-paused video playing, they do not take quality fluctuation into account. As a result, users may still suffer from the frequent fluctuation of video quality.

In the widely-used video streaming server DSS, quality control is implemented by monitoring packet delay (the difference between current time and the playing time of this packet). When a new packet is ready to be sent to streaming buffer, DSS checks the delay of it and compares this delay with some pre-set thresholds to judge whether and how violently the quality change should be made. This algorithm can reduce quality fluctuation to some degree. However, it only provides conservative quality change, which makes its reaction so slow that sometimes media player may have already stopped due to lack of video data before quality drops. Furthermore, this algorithm does not keep track of the bandwidth change so it can only adjust the video quality based on current state. Quality oscillation might occur in this algorithm.

The following of this paper will present our solution for each of the problems. The solutions then work together and serve as a solid foundation for a practical adaptive video streaming system. It should be noted that, in practical streaming systems another much concerned issue is the occasional packet loss during transmission, which will cause decoding errors and corrupted images, resulting in an significant degradation of video quality. How to avoid or even recover from packet loss/error is an important research topic that is popular in both networking and multimedia community. However, in this paper which focuses on server-side adaptation, packet loss is not handled and the video stream is assumed to be successfully delivered to the client.

@@

Being able to adjust bitrate is the prerequisite of an adaptive video streaming system. For SVC videos, it means to extract a subset of packets from the whole bitstream. How to optimally choose this subset to achieve minimum distortion under the given bitrate is the key point of bitstream extraction problem. In this section, we propose a linear error mode and a greedy-like algorithm to solve this problem.

@@

In H.264/AVC, the decoding process involves a linear feature described in \cite{Winken08}. According to this feature, the main steps of decoding -- prediction, transform and quantization -- are proximately linear operations (neglecting any rounding, clipping, and deblocking filtering), and the whole process can be expressed in a matrix format as follows:
\begin{equation}
\label{eq:linear}
s = {\rm\bf M} \cdot s + {\rm\bf T} \cdot c + p \: ,
\end{equation}
It means, the reconstructed samples of a Group of Pictures (GOP), denoted by $s$, can be viewed as a linear combination of previously reconstructed samples in the GOP, the residuals, and a static predictor. In (\ref{eq:linear}), $c$ refers to the transform coefficients and $p$ is a static predictor; ${\rm\bf M}$ and ${\rm\bf T}$ are square matrices such that ${\rm\bf M} \cdot s$ gives the Motion Compensated Prediction (MCP) signal and ${\rm\bf T} \cdot c$ gives the residuals. The actual values of ${\rm\bf M}$ depend on the selected macroblock types, reference indices and motion vectors, whereas the actual values of ${\rm\bf T}$ depend on the chosen Quantization Parameter (QP).

We apply the above linear decoding feature to the case of quality scalability of SVC and get a Linear Error Model (LEM). According to (\ref{eq:linear}), for a bitstream just containing the base layer, we have:
\begin{equation}
\label{eq:base0}
s_{B} = {\rm\bf M} \cdot s_{B} + {\rm\bf T}_{B} \cdot c_{B} + p \: ,
\end{equation}
where the subscript $B$ indicates the base layer variables. In quality scalability, the enhancement is mainly the refinement of transform coefficients, so for a bitstream containing an enhancement packet set $I_1$, the reconstruction relationship can be expressed as:
\begin{equation}
\label{eq:subset1}
s_{I_1} = {\rm\bf M} \cdot s_{I_1} + {\rm\bf T}_{B} \cdot c_{B} + \Big( \sum_{i \in I_1}{\rm\bf T}_i \cdot c_i \Big) + p \: ,
\end{equation}
where $i$ refers to an enhancement packet in the set $I_1$. $c_i$ is a vector containing coefficients within the enhancement packet $i$. ${\rm\bf T_i}$ is the transform matrix depending on the QP value of $i$. ${\rm\bf T}_i \cdot c_i$ gives the residual values decoded from $i$.

@@

Suppose $I_1$ is a subset of $I_2$, their reconstruction difference can be obtained by (\ref{eq:subset2}) - (\ref{eq:subset1}):
\begin{equation}
\label{eq:minus1}
(s_{I_2} - s_{I_1}) = {\rm\bf M} \cdot (s_{I_2} - s_{I_1}) + \Big( \sum_{i \in I_2}{\rm\bf T}_i \cdot c_i - \sum_{i \in I_1}{\rm\bf T}_i \cdot c_i \Big) \: ,
\end{equation}
\begin{equation}
\label{eq:minus2}
\Rightarrow e = s_{I_2} - s_{I_1} = ({\rm\bf I - M})^{-1} \cdot \Big( \sum_{i \in I_2}{\rm\bf T}_i \cdot c_i - \sum_{i \in I_1}{\rm\bf T}_i \cdot c_i \Big) \: .
\end{equation}

>>>>

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}
To cope with the challenges brought about by bandwidth fluctuation and improve the experience of watching online videos, an adaptive video streaming system that can adjust video quality according to actual network conditions is proposed based on the Scalable Video Coding (SVC) extension of H.264/AVC. First, a simple and effective linear error model is proposed and verified for quality scalability of SVC. The model exploits the linear feature of pixel value errors and can be used to accurately estimate the distortion caused by discarding any combination of enhancement data packets in an SVC bitstream. On that basis, a greedy-like algorithm is designed to assign each data packet a priority value according to its Rate-Distortion (R-D) impact, thus enabling R-D optimized bitstream extraction under certain bitrate constraints. Finally, the Proportional-Integral-Derivative (PID) method is utilized to control the video quality adjustment and determine a suitable bitrate for transmission. By monitoring and predicting past, current and future bandwidth information, the PID-based quality control algorithm is able to reduce quality fluctuation while still preserving a high quality level. Experimental results show that compared with the baseline software, the proposed system that integrates the above algorithms can achieve much lower video quality fluctuation, with PSNR variance reduced from 1.24 to 0.69, and at the same time deliver higher video quality, with the PSNR average increased by 0.83 dB.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
\IEEEPARstart{V}{ideo} streaming applications are becoming more and more popular on the Internet. Examples of the most common usage scenarios are Video-on-Demand (VoD), live broadcasting, video-based online education, etc. In these applications, video data are constantly delivered to different clients over various transmission channels, and the client media player can begin to play the video before the entire file has been transmitted. Among other features, video streaming makes it easy to watch videos anywhere and anytime, which represents the concept of Ubiquitous Multimedia Access (UMA), a very common topic within the multimedia community \cite{Gualdi08}. Due to its convenience, video streaming also attracts a large number of users and has become an increasingly important way for people to consume video content \cite{Chen13}.

In video streaming, the network conditions significantly influence the user's watching experience. To provide better network conditions, some researchers choose to adopt the new Software Defined Networking (SDN) architecture \cite{Egilmez14}. Based on a successful SDN paradigm named OpenFlow, noticeable results are achieved. For instance, a solution for improving throughput in hybrid networks is proposed in \cite{Li14} and a demonstration of OpenFlow-controlled network orchestration for video manycast is presented in \cite{Xue15}. Although these studies have offered a promising way to optimize network conditions, traditional best-effort networks still dominate currently, and the variety of user network environments remains to be one of the biggest challenges for video streaming. A user may use different types of networks (Ethernet, Wi-Fi, Cellular, etc) and may also switch from one type to another on going (e.g., from Wi-Fi to Cellular). Even in the same network, the end-to-end bandwidth will change from time to time, influenced by the overall traffic. Within such an environment, bandwidth fluctuation is inevitable, and it would be difficult or impossible for the user to get best watching experience if the video is streamed at a fixed bitrate. To cope with this challenge, the video streaming system should be able to adjust the video's bitrate or quality according to the network condition. In other words, the video streaming system needs to be adaptive.

There are generally two options for enabling adaptive video streaming. The first option is called simulcast, where several independent bitstreams, each with different video quality, are stored at the server, and the client selects one of them according to the currently available bandwidth. Dynamic Adaptive Streaming over HTTP (DASH) \cite{DASH}, which is standardized by the Moving Picture Expert Group (MPEG), belongs to this option. The second option is to adopt the Scalable Video Coding (SVC) \cite{SVC} technology, an extension of the H.264/AVC video coding standard \cite{SVCOverview}, which allows scalability of video quality and bitrate with only one single bitstream. In SVC, a video bitstream is composed of small data packets called Network Abstract Layer Units (NALUs). These NALUs are separated into one base layer for basic video quality and multiple enhancement layers for video quality improvement. By discarding the NALUs of the enhancement layers, the bitstream can be extracted to provide the most appropriate video quality and bitrate as required.

Due to their easy deployment \cite{Bouten14}, simulcast systems are already used by many commercial video websites (e.g., YouTube \cite{YouTube}) to enable quality adaptation. However, the adaptive range and granularity that simulcast provides is limited to the number of streams (YouTube videos have 5 quality levels at most: 240p, 360p, 480p, 720p and 1080p), and the processing and storage of multiple video streams require large amounts of computation time and disk space. In order to provide a wide range of seamless quality adaptation at reasonable time and space cost, the adoption of SVC is necessary. An SVC video bitstream can be extracted at the fine granularity of one NALU (a very small partition of data), and performance analysis \cite{SVCPerformance} shows that SVC also has much higher coding efficiency than simulcast, where the same content is encoded multiple times. Therefore, adopting SVC in the video streaming framework is technically a better option and has attracted a lot of attention in the academia \cite{Chuah12}-\cite{Cicalo14}.

While these studies mostly focus on Quality of Service (QoS) optimization and resource allocation for specific networks, there are two key problems that need to be primarily addressed in the SVC adaptive video streaming system. First, as a prerequisite for bitrate adaption, the whole bitstream needs to be extracted so that sub-streams with various bitrates can be obtained. This is called the bitstream extraction problem. Optimal bitstream extraction algorithm aims to extract a sub-stream with the best video quality, or minimal distortion, under the constraint of a given bitrate. Second, when the bandwidth changes with time, it is necessary to make decisions about whether and how to adjust the video quality or bitrate. This problem is called quality control, which means controlling the video quality (or the bitrate, from the transmission point of view) to follow the change of the bandwidth. A good quality control algorithm should determine a suitable bitrate according to current network conditions, and maintain the video quality to be both smooth and at a high level.

In this paper, an efficient adaptive video streaming system based on SVC is proposed with the above two problems better solved. Apple Darwin Streaming Server (DSS) \cite{DSS}, which is open-sourced and has been widely used to stream multimedia content in practice, is chosen as the basis for the proposed system. The original DSS project is modified and SVC support is added to it for the purpose of both research and application. Novel algorithms for bitstream extraction and quality control are then designed and implemented to enable better quality of service for the user. The contributions of our work can be summarized as follows:
\begin{itemize}
\item First, a simple and effective Linear Error Model (LEM) is proposed and verified, which can be used to accurately estimate the distortion caused by discarding any combination of data packets from an SVC bitstream.
\item Second, utilizing distortion estimation with LEM, a greedy-like algorithm is designed to assign priority for each data packet according to its Rate-Distortion (R-D) impact, thus enabling optimized bitstream extraction.
\item Finally, for the quality control problem, a novel algorithm based on the Proportional-Integral-Derivative (PID) controller is proposed to replace the native one in DSS, and much lower quality fluctuation is achieved.
\end{itemize}

@@

The rest of this paper is organized as follows. In section \ref{sec:analysis}, we analyze the problems in detail and introduce some related work on them. Then in section \ref{sec:bitstream-extraction}, the linear error model and the solution for the bitstream extraction problem are presented. The PID-based quality control algorithm is proposed in section \ref{sec:quality-control}. Section \ref{sec:experiment} lists the experimental results and compares our proposed solution with others. Finally, in section \ref{sec:conclusion}, we conclude this paper and point out related future work.

@@

In this section, we analyze the two most important problems in SVC adaptive video streaming, i.e., bitstream extraction and quality control, and introduce some related work on them.

@@

Bitstream extraction is the operation of choosing a subset of data packets from a whole SVC bitstream to form a sub-stream. Fig. \ref{fig:Bitstream-Extraction} illustrates the structure of an SVC bitstream and a possible way of extracting a sub-stream from it. The data packets in the bitstream are divided into different layers. The temporal layer (T0$\sim$T3) reflects the reference or dependency relationship between the frames. The arrows ot the top point from the referenced frame to the frame which is dependent on it. The quality layer (Q0$\sim$Q2) indicates the scalability of the video quality within each frame. Data packets in the higher quality layers (i.e., Q1 and Q2) are partially discarded to implement the extraction. For example, the dashed line demonstrates the extraction which will keep the data packets below the line and discard all the packets above it.

@@

As we can see, in its nature SVC bitstream extraction is a combinatorial optimization problem. There are many combinations of packets to be discarded, and we want to choose a combination that maximizes the video quality of the extracted sub-stream, under the constraint of a given bitrate. More specifically, this problem is in the form of the ``0-1 Knapsack Problem" \cite{Knapsack}. Each data packet can be treated as an object with certain weight (the packet's size) and value (its contribution to the video quality). The solving process is mainly deciding whether or not a packet should be included in the final extracted sub-stream, or, in terms of the ``0-1 Knapsack Problem", choosing which object to pack up.

For a typical ``0-1 Knapsack Problem", the optimal solution can generally be found using the dynamic programming algorithm. However, this is not operational for the bitstream extraction problem, mainly due to the dependence between the data packets. First, the subset of packets to be extracted cannot be chosen arbitrarily, because if a packet is included, all the packets it depends on must also be included. For example, in Fig. \ref{fig:Bitstream-Extraction}, we cannot include the Q2 packet of a frame without including the Q1 packet of the same frame. Second, the value of a packet is not obvious and certain, because a packet's contribution to video quality depends on other packets and may vary for different extractions. There is no way to exactly determine the video quality of an extracted sub-stream, unless we perform the extraction and calculate the distortion for real. To the best of our knowledge, no method presented in the literature claims to find the optimal solution of the SVC bitstream extraction problem. All the efforts are made to obtain a solution that is as good as possible.

In \cite{Amonou07}, Amonou et al. proposed a bitstream extraction framework based on the concept of \textit{Quality Layers}. This framework has been adopted by the SVC reference software, i.e., JSVM \cite{JSVM}, and achieves better R-D performance compared to the basic extractor of JSVM. In Amonou's framework, a Quality Layer value, which reflects a data packet's R-D impact, is calculated and used for R-D optimized bitstream extraction. The process of obtaining the distortion impact of every packet is extremely time consuming, and the distortion estimation accuracy can also be further improved. Therefore, other distortion/error models were proposed to either reduce the computational complexity or improve the estimation accuracy. Sun et al. \cite{Sun09} and Maani et al. \cite{Maani09} constructed models to calculate the distortion based on the drift propagation between frames. Sun et al.'s method is almost as good as JSVM in terms of performance, but has much reduced complexity. Maani et al.'s model can achieve better estimation accuracy than JSVM; however, since it is a training-based model, more computation is needed to get robust model parameters. Ramanathan et al. \cite{Ramanathan12} proposed their concept and calculation method of the ``Quality Metadata'' to signal and use in bitstream extraction. Yang et al. \cite{Yang13} proposed a method based on simulated annealing to solve the bitstream extraction problem. In both methods, the accuracy of distortion estimation affects the performance significantly. With an accurate estimation of a packet's distortion impact, its prioritization order during extraction can be determined. A priority value can then be assigned to it and stored in the bitstream for optimized extraction in the future. In \cite{Lim06}, Lim et al. treated the SVC bitstream as a tree structure and assigned extraction priority accordingly. In \cite{Zhu11}, Zhu et al. chose to assign priority values derived using the Langrange multiplier method.

In our previous work \cite{Zhang12}, we proposed a linear error model for distortion estimation and priority assignment, based on which good bitstream extraction performance is achieved. However, this work assumes that the original video sequence is available for reference, which is not always true. How to handle the special characteristic of the no-reference case is lacking in the previous work and included in this paper.

@@

Quality control is the second problem we will address in this paper. A quality control algorithm gives the suitable bitrate for transmission, which is directly related to the change of the video quality delivered to the user. In an SVC streaming system, the bitrate determined by the quality control algorithm is used as the constraint under which bitstream extraction is performed. Even in non-SVC streaming systems (DASH, for example), quality control is an important aspect that requires careful design.

For a quality control algorithm, there are three targets to be achieved. The first is ensuring the continuity of video play. In real-time streaming systems, each data packet has its own playtime, which is also treated as the deadline of the transmission. If the packet does not arrive at the client side before the deadline, the media player will be paused due to lack of data, and that should be avoided. The second is maintaining smooth video quality. Although network conditions may change rapidly, it is necessary for the streamed video to maintain a relatively stable quality level, because frequent change of video quality will lead to poor user experience. The third target is to make the video quality as good as possible. In other words, the available bandwidth should be fully utilized. Suppose we keep sending video data at a low bitrate (with low video quality), neither will the playback pause nor will any quality fluctuation occur. However, a low video quality is certainly not what the user wants.

The state-of-the-art solutions for avoiding playback pause in the client media player mainly focus on how to select and schedule the video packets. In \cite{Gao06}, K. Gao et al. proposed a priority-based scheduling strategy, which guarantees that packets of higher importance will be transmitted earlier. Among the packets with the same priority, those with the earliest deadline will be transmitted first. A similar solution is proposed in \cite{Schierl10}, where the video data are pre-buffered in several buffers of different priorities. Whenever re-buffering occurs, the quality will be degraded to a lower level to meet the time and rate constraints. Although the priority-based scheduling strategies perform well in maintaining a non-paused video play, they do not take quality fluctuation into account. As a result, users may still suffer from the frequent fluctuation of video quality.

In the widely used video streaming server DSS, quality control is implemented by monitoring packet delay (the difference between the current time and the playing time of this packet). When a new packet is ready to be sent to the streaming buffer, DSS checks its delay and compares this delay with some pre-set thresholds to judge whether a quality change should be made and how extreme it should be. This algorithm can reduce quality fluctuation to some degree. However, it only provides conservative quality changes, which makes its reaction so slow that sometimes the media player may have already stopped due to lack of video data before the quality drops. Furthermore, this algorithm does not keep track of the bandwidth change, and therefore, it can only adjust the video quality based on the current state. Quality oscillation might occur in this algorithm.

In the following part of this paper, we will present our solution for each of the problems. The solutions then work in combination and serve as a solid foundation for a practical adaptive video streaming system. It should be noted that in practical streaming systems, another issue of concern is the occasional packet loss during transmission, which will cause decoding errors and corrupted images, resulting in a significant degradation of video quality. Avoiding or even recovering from packet loss/error is an important research topic that is popular in both the networking and multimedia communities. However, in this paper, which focuses on server-side adaptation, packet loss is not handled, and the video stream is assumed to be successfully delivered to the client.

@@

Being able to adjust the bitrate is the prerequisite of an adaptive video streaming system. For SVC videos, this means extracting a subset of packets from the whole bitstream. How to optimally choose this subset to achieve minimum distortion under the given bitrate is the key point of the bitstream extraction problem. In this section, we propose a linear error model and a greedy-like algorithm to solve this problem.

@@

In H.264/AVC, the decoding process involves a linear feature described in \cite{Winken08}. According to this feature, the main steps of decoding -- prediction, transform and quantization -- are approximately linear operations (neglecting any rounding, clipping, and deblocking filtering), and the whole process can be expressed in a matrix format as follows:
\begin{equation}
\label{eq:linear}
s = {\rm\bf M} \cdot s + {\rm\bf T} \cdot c + p \: .
\end{equation}
It means, the reconstructed samples of a Group of Pictures (GOP), denoted by $s$, can be viewed as a linear combination of previously reconstructed samples in the GOP, the residuals, and a static predictor. In (\ref{eq:linear}), $c$ refers to the transform coefficients, and $p$ is a static predictor; ${\rm\bf M}$ and ${\rm\bf T}$ are square matrices such that ${\rm\bf M} \cdot s$ gives the Motion Compensated Prediction (MCP) signal and ${\rm\bf T} \cdot c$ gives the residuals. The actual values of ${\rm\bf M}$ depend on the selected macroblock types, reference indices and motion vectors, whereas the actual values of ${\rm\bf T}$ depend on the chosen Quantization Parameter (QP).

We apply the above linear decoding feature to the case of quality scalability of SVC and obtain a Linear Error Model (LEM). According to (\ref{eq:linear}), for a bitstream just containing the base layer, we have:
\begin{equation}
\label{eq:base0}
s_{B} = {\rm\bf M} \cdot s_{B} + {\rm\bf T}_{B} \cdot c_{B} + p \: ,
\end{equation}
where the subscript $B$ indicates the base layer variables. In quality scalability, the enhancement is mainly the refinement of the transform coefficients. For a bitstream containing an enhancement packet set $I_1$, the reconstruction relationship can be expressed as:
\begin{equation}
\label{eq:subset1}
s_{I_1} = {\rm\bf M} \cdot s_{I_1} + {\rm\bf T}_{B} \cdot c_{B} + \Big( \sum_{i \in I_1}{\rm\bf T}_i \cdot c_i \Big) + p \: ,
\end{equation}
where $i$ refers to an enhancement packet in the set $I_1$, $c_i$ is a vector containing coefficients within the enhancement packet $i$, ${\rm\bf T_i}$ is the transform matrix depending on the QP value of $i$ and ${\rm\bf T}_i \cdot c_i$ gives the residual values decoded from $i$.

@@

If $I_1$ is a subset of $I_2$, their reconstruction difference can be obtained by (\ref{eq:subset2}) - (\ref{eq:subset1}):
\begin{equation}
\label{eq:minus1}
(s_{I_2} - s_{I_1}) = {\rm\bf M} \cdot (s_{I_2} - s_{I_1}) + \Big( \sum_{i \in I_2}{\rm\bf T}_i \cdot c_i - \sum_{i \in I_1}{\rm\bf T}_i \cdot c_i \Big) \: ,
\end{equation}
\begin{equation}
\label{eq:minus2}
\Rightarrow e = s_{I_2} - s_{I_1} = ({\rm\bf I - M})^{-1} \cdot \Big( \sum_{i \in I_2}{\rm\bf T}_i \cdot c_i - \sum_{i \in I_1}{\rm\bf T}_i \cdot c_i \Big) \: .
\end{equation}

In particular, if the difference between $I_2$ and $I_1$ is a single enhancement packet $j$, i.e., $I_2 \setminus I_1 = \{j\}$, then according to (\ref{eq:minus2}), the difference between the two reconstruction vectors can be written as
\begin{equation}
\label{eq:error_j}
e_j = ({\rm\bf I - M})^{-1} \cdot {\rm\bf T}_j \cdot c_j \: .
\end{equation}
